---
title: 'Statistical Methods for Criminology: Practical Four'
author: "Ben Matthews"
date: "2024-09-04"
output: github_document
---

# getting started ---------------------------------------------------------

 the following lines will install packages on your computer if you don't have them already

```{r}

if(!require(tidyverse)){
  install.packages("tidyverse")
  library(tidyverse)
}

if(!require(broom)){
  install.packages("broom")
  library(broom)
}

if(!require(clarify)){
  install.packages("clarify")
  library(clarify)
}

```

# data --------------------------------------------------------------------



 see pages 58-63 at https://www.gov.scot/binaries/content/documents/govscot/publications/statistics/2020/09/simd-2020-technical-notes/documents/simd-2020-technical-notes/simd-2020-technical-notes/govscot%3Adocument/SIMD%2B2020%2Btechnical%2Bnotes.pdf
 because counts smaller than three were suppressed due to disclosure control, i have simulated crime counts
 for datazones with fewer than three crimes, allocating them 0, 1 or 2 crimes with equal probability
 i also created a simulated violent crime variable by multiplying the total crime count for the
 datazone by 0.24 (as non-sexual crimes of violence make up 24% of recorded crime
 see # from https://www.gov.scot/publications/recorded-crime-scotland-2023-24/pages/4/)
 these variables are just intended to illustrate the methods discussed in this course - they won't tell us
 anything informative about crime in Scotland (unless by accident)
 all other data comes from the Scottish Index of Multiple Deprivation ranks and indicators


# read in the data --------------------------------------------------------

```{r}

simd <- readRDS(url("https://github.com/benmatthewsed/statistical-methods-criminology-slides/raw/master/resources/simd_crime_sim.rds"))

```

# let's fit the model from the first practical again

```{r}
qp_mod <- 
  glm(crime_integer_sim ~ simd2020_rank,
      family = "quasipoisson",
      data = simd)

```

 we can use this model to visualize predicted values of expected crime at different
 levels of SIMD

```{r}

simd_range <- data.frame(simd2020_rank = seq(1, 6976)) # make a prediction for all SIMD ranks

data.frame(
  predicted_crime = 
    predict(qp_mod,
        newdata = simd_range, # make a prediction for all SIMD ranks
        type = "response") # gives us predictions on the response scale
  ) |> 
  mutate(simd2020_rank = seq(1, 6976)) |> 
  ggplot(aes(x = simd2020_rank, y = predicted_crime)) +
  geom_line()

```


# plotting predictions with two variables ---------------------------------


 maybe the overall relationship between deprivation and crime works differently in cities versus other areas?
 we'll compare SIMD and recorded crime in 'City' local authorities to other local authorities
 (note that we could have used other operationalizations here, such as the Scottish Government's rural/urban lassification)
 create our city_la variable based on whether the LA name has 'City' in it

```{r}

simd <- 
simd |> 
  mutate(city_la = if_else(str_detect(council_area, "City"),
                           1,
                           0))

```

 fit our count model, including an interaction between SIMD and city_la
 so city LAs get their own SIMD trend, and so do non-city LAs

 (without the interaction the trends are forced to be parallel)

```{r}

qp_mod <- 
glm(crime_integer_sim ~ simd2020_rank * city_la,
    family = "quasipoisson",
    data = simd)

```

 let's check the coefficients

```{r}

qp_mod |> 
  broom::tidy()

```

 the main effect of city la is not statistically significant at p < 0.05, but
 the interaction is
 therefore we keep both

 we can calculate adjusted predictions to visualize this interaction

 we can do this just using predict(), asking it to give us predictions for all
 datazone SIMD ranks (1 to 6976), and under the condition of being in a
 city_la or not

```{r}

pred_df <- tidyr::crossing(simd2020_rank = seq(1:6976),
                           city_la = c(0, 1))

simd_preds <- 
predict(
  qp_mod,
  newdata = pred_df,
  type = "response"
)

pred_df |> 
  mutate(pred = simd_preds) |> 
  ggplot(aes(x = simd2020_rank, y = pred, colour = factor(city_la))) +
  geom_line()

```

 the lines are very similar where simd2020_rank is close to zero
 (which explains why the main effect was not statistically significant)
 but they diverge as SIMD rank increases
 deprived non-city neighbourhoods are quite similar to deprived city neighourhoods
 but in places with low deprivation city neighbourhoods and non-city neighbourhoods
 seem quite different


 this is fine so far, but really we want to assess the uncertainty in these estimates

# we can do this with {clarify}

 first we simulate coefficients from our model

```{r}

sims <- clarify::sim(qp_mod,
                     n = 1000) # this is the default number.  try increasing it and see what happens

```

 we can use sim_setx to make predictions for the outcome (on the outcome scale)
 across a grid of predictors that we provide

```{r}

sim_res <- sim_setx(sims, x = list(simd2020_rank = seq(1:6976), city_la = c(1, 0)),
         verbose = TRUE) # verbose = TRUE gives a status bar showing how long the simulation is taking

plot(sim_res)

```

 the solid lines are the same as those we calculated by hand, but we now get error bands
 showing the 95% confidence intervals
 these bands overlap at high levels of deprivation, showing that - based on our model - 
 it looks like city and non-city areas that are very deprived have similar crime rates




# questions ---------------------------------------------------------------

 based on the description of the SIMD crime data at
 see pages 58-63 at https://www.gov.scot/binaries/content/documents/govscot/publications/statistics/2020/09/sim-2020-technical-notes/documents/simd-2020-technical-notes/simd-2020-technical-notes/govscot%3Adocument/SIMD%2B220%2Btechnical%2Bnotes.pdf
 and the discussion in sessions three and four
 write a brief description of the figure produced by the code above, and reflect on the value
 of the prediction and simulation approach to translating the coefficients

 what are the benefits? what are the limitations?



# going further: using clarify with our own function -------------------------------------

 in the last example we used a built in function from clarify
 now we'll recreate the example from the lectures of calculating
 victimization inequality


 load in the data

```{r}
dat <-
  tribble(
    ~prev, ~year, ~sex, ~n,
    0.167, "2015", "men", 15030,
    0.153, "2015", "women", 18320,
    0.197, "2020", "men", 15505,
    0.189, "2020", "women", 18230
  )

```

 calculate the number of victims

```{r}

dat <-
  dat |> 
  mutate(vict = as.integer(n * prev))

```

 define our victim divide function


```{r}
victim_divide <- function(base_y1, base_y2){
  ((base_y2 - 1) - (base_y1 - 1)) / (base_y1 - 1)
}

```

 fit a model to calculate the odds of being a victim for men and women,
 with an interaction with year
 for clarify we need the results to be part of the same model - in the example in the slides
 we fit two seperate models, to dollow the format of Hunter and Tseloni (2016)

```{r}

mod1 <- glm(cbind(vict, n - vict) ~ fct_rev(sex) * year,
            family = "binomial",
            data = dat)

s <- sim(mod1,
         n = 1000)

```

 this function is quite complicated, but it's just extracting the right coefficients

```{r}

sim_victim_divide <- function(coefs) {
  
  men_2015 <- unname(coefs["fct_rev(sex)men"])
  men_interact <- unname(coefs["fct_rev(sex)men:year2020"])
  
  men_2020 <- men_2015 + men_interact
  
  or_2015 <- exp(men_2015)
  or_2020 <- exp(men_2020)
  
  victim_divide(or_2015, or_2020)
  
}

```

 test on our model

```{r}

sim_victim_divide(coef(mod1))

```

now apply


```{r}
est2 <- sim_apply(s, 
                  FUN = sim_victim_divide)

tibble(x = as.vector(est2)) |> 
  reframe(vds = quantile(x, c(0.01, 0.025, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.975, 0.99)),
          vals = c(0.01, 0.025, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.975, 0.99))

```

 as previously, the 97.5 percenticle is above 0, suggesting that the change in the VD over time
 is not statistically significant
 note that the median VD from our simulations is not identical to the result we calculated
 from the original model coefficients
 this is called monte-carlo error - and as a result, the point estimates from the original model
 are preferable to the median of the simulations
 see Rainey, C. (2023). A careful consideration of CLARIFY: Simulation-induced bias in point estimates of quantities of interest. Political Science Research and Methods, 1–10. https://doi.org/10.1017/psrm.2023.8



# going even further ------------------------------------------------------

 read the clarify documentation at https://iqss.github.io/clarify/
 fit another model to the SIMD data and desribe the results using some of clarify's
 functionality
