---
title: 'Statistical Methods for Criminology: Practical One'
author: "Ben Matthews"
date: "2024-09-04"
output: github_document
---

# getting started ---------------------------------------------------------

the following lines will install packages on your computer if you don't have them already

```{r}


if(!require(tidyverse)){
  install.packages("tidyverse")
  library(tidyverse)
}

if(!require(broom)){
  install.packages("broom")
  library(broom)
}

```

# data --------------------------------------------------------------------



 see pages 58-63 at https://www.gov.scot/binaries/content/documents/govscot/publications/statistics/2020/09/simd-2020-technical-notes/documents/simd-2020-technical-notes/simd-2020-technical-notes/govscot%3Adocument/SIMD%2B2020%2Btechnical%2Bnotes.pdf

 because counts smaller than three were suppressed due to disclosure control, i have simulated crime counts
 for datazones with fewer than three crimes, allocating them 0, 1 or 2 crimes with equal probability

 i also created a simulated violent crime variable by multiplying the total crime count for the
 datazone by 0.24 (as non-sexual crimes of violence make up 24% of recorded crime
 see # from https://www.gov.scot/publications/recorded-crime-scotland-2023-24/pages/4/)
 these variables are just intended to illustrate the methods discussed in this course - they won't tell us
 anything informative about crime in Scotland (unless by accident)

 all other data comes from the Scottish Index of Multiple Deprivation ranks and indicators


# read in the data --------------------------------------------------------

```{r}

simd <- readRDS(url("https://github.com/benmatthewsed/statistical-methods-criminology-slides/raw/master/resources/simd_crime_sim.rds"))
```

# descriptives ------------------------------------------------------------

 our key outcome is crime_integer_sim

```{r}

simd |> 
  ggplot(aes(x = crime_integer_sim)) +
  geom_histogram(binwidth = 1)

```

 very skewed!
 the average crime count is just over 22
 but there is a lot of variability around this (sd of 32), and no datazones with less than zero crimes
 ... which makes sense!

```{r}

simd |> 
  summarise(mean_crime = mean(crime_integer_sim),
            sd_crime = sd(crime_integer_sim),
            min_crime = min(crime_integer_sim),
            max_crime = max(crime_integer_sim))

```
 lets do the same for violent crime


```{r}
simd |> 
  summarise(mean_vio = mean(vio_integer_sim),
            sd_vio = sd(vio_integer_sim),
            min_vio = min(vio_integer_sim),
            max_vio = max(vio_integer_sim))

```


 we will use overall SIMD (simd2020v2_rank) as our main independent variable
 this ranks all 6976 datazones from most (1) to least (6976) deprived
 so it is uniformly distributed with a mean halfway from 1 to 6976

```{r}

simd |> 
  summarise(mean_simd = mean(simd2020_rank),
            sd_simd = sd(simd2020_rank),
            min_simd = min(simd2020_rank),
            max_simd = max(simd2020_rank))

```



# regression modelling ----------------------------------------------------

 we will compare linear and count models for overall crime and violence

# linear model

```{r}

lm(crime_integer_sim ~ 1, data = simd) |> 
  broom::tidy()

```

 the intercept is the same as the mean crime count we calculated manually - that's good!

```{r}

glm(crime_integer_sim ~ 1, 
    family = "poisson",
    data = simd) |> 
  broom::tidy()

```

 the estimate is 3.10. we need to exp() this to get the estimate on the outcome scale
 this is because the poisson model uses the log link function to link the coefficients
 to the observed data

```{r}

glm(crime_integer_sim ~ 1, 
    family = "poisson",
    data = simd) |> 
  broom::tidy() |> 
  mutate(exp_est = exp(estimate))

```
 exp_est is 22.3 - same as that we calculated already

# now let's try a quasipoisson


```{r}
glm(crime_integer_sim ~ 1, 
    family = "quasipoisson",
    data = simd) |> 
  broom::tidy() |> 
  mutate(exp_est = exp(estimate))

```
 the results are the same! but note the std.error column - this is now
 much larger (0.0176) than in the previous model (0.00254)
 this reflects the quasi poisson correcting for over-dispersion


# crime and SIMD ----------------------------------------------------------

 let's look at the relationship between crime and SIMD

```{r}

mod_crime_simd_lm <- lm(crime_integer_sim ~ simd2020_rank, data = simd)

mod_crime_simd_lm |> 
  broom::tidy()

```

 SIMD is negative and statistically significant - with less deprivation (i.e. as SIMD increases)
 we would expect less crime

```{r}
mod_crime_simd_p <- glm(crime_integer_sim ~ simd2020_rank, family = "poisson", data = simd)

mod_crime_simd_p |> 
  broom::tidy() |> 
  mutate(exp_est = exp(estimate))

```
 the poisson model tells us the same, although the exp_est is rounded to 1.00
 if we calculate this directly we see it is 0.9997

```{r}
exp(-0.000271)

```

 for every increase in SIMD you have 99.97% as much crime, or if you like, a 
 1 - exp(-0.000271) - 0.027% increase in crime



# modelling violence ------------------------------------------------------

```{r}
mod_vio_simd_lm <- lm(vio_integer_sim ~ simd2020_rank, 
                      data = simd)

mod_vio_simd_p <- glm(vio_integer_sim ~ simd2020_rank, 
                      family = "poisson",
                      data = simd)

mod_vio_simd_lm |> 
  broom::tidy() |> 
  mutate(conf_low = estimate - 1.96 * std.error, # we can calculate confidence intervals for these coefficients
         conf_upp = estimate + 1.96 * std.error)

```

 as before, simd is negatively and statistically significantly related to our crime indicator

# let's see this in the poisson model

```{r}

mod_vio_simd_p |> 
  broom::tidy() |> 
  mutate(exp_est = exp(estimate),
         conf_low = estimate - 1.96 * std.error,
         conf_upp = estimate + 1.96 * std.error,
         exp_conf_low = exp(conf_low),
         exp_conf_upp = exp(conf_upp))

```

 we might want to look at a quasipoisson model in case the the standard errors in the poisson
 are too small

```{r}

mod_vio_simd_qp <- glm(vio_integer_sim ~ simd2020_rank, 
                      family = "quasipoisson",
                      data = simd)

mod_vio_simd_qp |> 
  broom::tidy() |> 
  mutate(exp_est = exp(estimate),
         conf_low = estimate - 1.96 * std.error,
         conf_upp = estimate + 1.96 * std.error,
         exp_conf_low = exp(conf_low),
         exp_conf_upp = exp(conf_upp))
```
 the standard error is much larger - this might indicate that we should be worried about over-dispersion


# over to you -------------------------------------------------------------

 using either the models above, or by using the additional variables in the SIMD data to fit your own model,
 write a summary of what this analysis can and can't tell us about the relationship between deprivation
 and crime in Scotland
 what are the pros and cons of the different modelling approaches?
 what are the limitations of the data used?



